Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway

### 2022-02-19 analysis/paper/paper.Rmd
Minor edits

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research, which has been based on more subjectively driven methods. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, several artefact categories have here been collapsed for the CA (Figure \@ref(fig:variable-aggregation)). Artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading. An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

# A reviewer wanted artefact types combined irrespective of raw-material and 
# non-flint as supplementary column. I start by including it as a normal 
# column.
all_nonflint_prop <- sites %>%
    mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined + 
      projectile_nonflint,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch +
      knife_nonflint + scraper_nonflint + drill_nonflint + 
      microblade_retouch_nonflint + blade_retouch_nonflint + 
      fragment_retouch_nonflint +
      flake_retouch_nonflint,
    macro_tool = axe + chisel + axe_nonflint + 
      chisel_nonflint + club_hatchet_nonflint,
    core = core + core_nonflint,
    core_fragment = core_fragment + core_fragment_nonflint,
    axe_debitage = axe_debitage + axe_debitage_nonflint,
    blade = blade + blade_nonflint,
    microblade = microblade + microblade_nonflint,
    fragment = fragment + fragment_nonflint,
    flake = flake + flake_nonflint,
    chip = chip + chip_nonflint,
    nonflint_prop = nonflint_count / artefact_count,
    .keep = "unused") %>% 
    # Make site names the rownames
    column_to_rownames(var = "no") %>% 
    # Deselect a few of the columns not be used
    select(-excavation_project, -excavated_m3, -dating_method)

@@

# Version of the above where the proportion of non-flint is simply added as a 
# supplementary column, following suggestion from the reviewer. 
moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
                       -reported_start_bc, -reported_end_bc,
                       -askeladden_id)

>>>>

---
title: "Exploring the Composition of Lithic Assemblages in Mesolithic South-Eastern Norway"
header-includes:
   - \usepackage{lineno}
   - \usepackage{afterpage}
   - \usepackage{fancyhdr}
   - \linenumbers
author:
  - Isak Roalkvam:
      email: isak.roalkvam@iakh.uio.no
      institute: [UiO]
      correspondence: false
institute:
  - UiO: University of Oslo, Department of Archaeology, Conservation and History
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::pdf_document2:
      toc: false
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/saa.csl" # Insert path for the bib-style
abstract: |
  This paper leverages multivariate statistics to explore the composition of 54 Mesolithic assemblages located in south-eastern Norway. To provide analytical control pertaining to factors such as variable excavation practices, systems for artefact categorisation and raw-material availability, the sites chosen for analysis have all been excavated relatively recently and have a constrained geographical distribution. The assemblages were explored following two strains of analysis. The first of these entailed the use of artefact categories that are in established use within Norwegian Mesolithic archaeology, while the other involved drawing on measures that have been linked directly to land-use and mobility patterns associated with lithic assemblages more widely. The findings pertaining to the established artefact categories largely reflect the temporal development previously reported in Norwegian Mesolithic research. Furthermore, the chronological trends associated with variables taken from the so-called Whole Assemblage Behavioural Indicators (e.g. Clark and Barton 2017), originally devised for characterising Palaeolithic assemblages in terms of associated mobility patterns, also align with the development previously proposed in the literature. This provides an initial indication that these measures are applicable in a Norwegian Mesolithic setting as well, setting the stage for a more targeted and rigorous model evaluation outside this exploratory setting. Furthermore, this finding supports the notion that these measures can offer a powerful comparative tool in the analysis of lithic assemblages more generally.

@@

The aim of the first of part of the analysis conducted here is to evaluate the degree to which the composition of the assemblages align with earlier studies that have employed more informal methods. This therefore assumes that the artefact categories employed in Norwegian Stone Age archaeology are, at least to a certain extent, behaviourally meaningful. However, the approach taken is also partially informed by the so-called Frison effect [@jelinek1976], which pertains to the fact that lithics studied by archaeologists can have had long and complex use-lives in which they took on a multitude of different shapes before they were ultimately discarded. Several scholars have built on this to argue that morphological variation in retouched lithics from the Palaeolithic cannot be assumed to predominantly be the result of the intention of the original knapper to reach some desired end-product, but rather that what is commonly categorised as discrete types of artefacts by archaeologists can instead in large part be related to variable degrees of modification through use and rejuvenation [e.g. @barton1991; @dibble1995; @barton2021]. Consequently, artefact categories believed not to be internally consistent and categorically exclusive have therefore been collapsed for the analysis, as their contribution as discrete analytical units could potentially be misleading (Figure \@ref(fig:variable-aggregation)). An underlying assumption of the largely intuitively determined aggregation procedure is therefore effectively that the retained categories represent artefact categories that have fulfilled different purposes or are related to different technological processes. While aggregating artefact categories in this manner could subsume important variation, it does also reduce the possibility that any conclusions are not simply the result of employing erroneous units of analysis.

@@

# Based on suggestion from reviewer: artefact types combined irrespective of 
# raw-material and  non-flint as supplementary column. I start by including 
# it as a normal column.
all_nonflint_prop <- sites %>%
    mutate(
    projectile = microlith + tanged_point + transverse_point + 
    single_edged_point + hognipen_point + projectile_undefined + 
      projectile_nonflint,
    small_tool = scraper + knife + burin + drill + blade_retouch + 
      microblade_retouch + chip_retouch + fragment_retouch + flake_retouch +
      knife_nonflint + scraper_nonflint + drill_nonflint + 
      microblade_retouch_nonflint + blade_retouch_nonflint + 
      fragment_retouch_nonflint +
      flake_retouch_nonflint,
    macro_tool = axe + chisel + axe_nonflint + 
      chisel_nonflint + club_hatchet_nonflint,
    core = core + core_nonflint,
    core_fragment = core_fragment + core_fragment_nonflint,
    axe_debitage = axe_debitage + axe_debitage_nonflint,
    blade = blade + blade_nonflint,
    microblade = microblade + microblade_nonflint,
    fragment = fragment + fragment_nonflint,
    flake = flake + flake_nonflint,
    chip = chip + chip_nonflint,
    nonflint_prop = nonflint_count / artefact_count,
    .keep = "unused") %>% 
    # Make site names the rownames
    column_to_rownames(var = "no") %>% 
    # Deselect a few of the columns not be used
    select(-excavation_project, -excavated_m3, -dating_method)

@@

# Version of the above where the proportion of non-flint is simply added as a 
# supplementary column, following suggestion from reviewer. 
moddat <- select(all_nonflint_prop, -site, -mean_date, -flint_count, 
                       -reported_start_bc, -reported_end_bc,
                       -askeladden_id)
